\textbf{Adaptation of the PHQ-4 Short Screening for Depression and Anxiety to increase its Sensitivity to Subclinical Variability}

### 2023-01-09 paper/manuscript.Rmd
update

keywords          : "PHQ-4, depression, anxiety, brief questionnaire, short scale"
wordcount         : "1011"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

The added value in terms of finer sensitivity that we hypothesize should be confirmed by cross-validating the PHQ-4R with more comprehensive measures of anxiety and depression. Additionally, evidence comparing the refined version against the original PHQ-4 would be useful to better understand whether the new "Once or twice" option pulls participants that would have otherwise answer "Not at all" or "Several days". Finally, future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations) to see if it affects the cut-offs proposed in the literature (and whether it allows for a better discrimination of mood disorders), and refine if need be the normalized scoring estimates in larger sample.

@@

```{r message=FALSE, warning=FALSE}
# Load the function
source("https://raw.githubusercontent.com/DominiqueMakowski/PHQ4R/main/score_PHQ4.R")

score_PHQ4(A1 = "Not at all", A2 = "Once or twice",
           D1 = "Once or twice", D2 = "several days",
           method = "basic")
```

>>>>

keywords          : "PHQ-4, depression, anxiety, brief questionnaire, short scale"
wordcount         : "1158"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

The hypothesized added value in terms of finer sensitivity should be confirmed by cross-validating the PHQ-4R with more comprehensive measures of anxiety and depression. Additionally, evidence comparing the refined version against the original PHQ-4 would be useful to better understand whether the new "Once or twice" option pulls participants that would have otherwise answer "Not at all" or "Several days". Finally, future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations) to see if it affects the cut-offs proposed in the literature (and whether it allows for a better discrimination of mood disorders), and refine if need be the normalized scoring estimates in larger sample.

@@

```{r echo=TRUE, message=FALSE, warning=FALSE}
# Load the function
source("https://raw.githubusercontent.com/DominiqueMakowski/PHQ4R/main/score_PHQ4.R")

# Enter the responses
score_PHQ4(A1 = "Not at all", A2 = "Once or twice",
           D1 = "Once or twice", D2 = "several days",
           method = "basic")
```

### 2023-01-09 paper/manuscript.tex
update

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{PHQ-4, depression, anxiety, brief questionnaire, short scale\newline\indent Word count: 1011}
\usepackage{lineno}

@@

\abstract{%
The PHQ-4 is an ultra brief (4 items) screening questionnaire for depression and anxiety. We propose to add one aditional response option (``Once or twice'', in between ``Not at all'' and ``Several days'') to improve its sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. Using Item Response Theory (IRT), we show that this new option does indeed capture specific portions of the measured constructs, and provide scoring procedure that preserve the compatability with past versions.

\textbf{Significance Statement}. Developing reliable and sensitive instruments for mood disorders screening is critical in a global context marked by international crises (pandemics, wars), where more and more surveys as done online. In this study, we show that a small modification to the widely used PHQ-4 scale (adding the ``Once or twice'' response option) can increase its ability to capture the mild fluctuations prevalent in subclinical samples.
}

@@

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety (Kroenke et al., 2009). It consists of two items for depression (PHQ--2, Kroenke et al., 2003) and anxiety (GAD--2, Kroenke et al., 2007), each corresponding to DSM-V Diagnostic Criterion A symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations (Christodoulaki et al., 2022; Materu et al., 2020; Mendoza et al., 2022), becoming one of the most popular screening instruments for depression and anxiety (Maurer et al., 2018).

@@

\hypertarget{original-scale}{%
\subsection{Original Scale}\label{original-scale}}

@@

\hypertarget{revision}{%
\subsection{Revision}\label{revision}}

@@

The sample consists of 497 English-speaking participants (Mean age = 30.0, SD = 10.1, range: {[}18, 73{]}; Sex: 49.9\% females, 50.1\% males) from the general population who were administered the refined PHQ-4 online as part of another study, which data is available in open-access at \url{https://github.com/RealityBending/IllusionGameReliability}.

This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187). All participants provided their informed consent prior to participation and were incentivized after completing the study.

@@

\hypertarget{descriptive}{%
\subsection{Descriptive}\label{descriptive}}

@@

The objective of this study was to test the introduction of a ``Once or twice'' response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. The fact that this new response option was the most prevalent response made by participants in our study is in itself evidence for its usefulness. IRT analysis further revealed that this response captures with precision a unique portion of anxiety and depression. Taken together, our results suggest that adding this option response increases the scale's potential to discriminate average mood levels (which are superior to zero) from lower-end extremes (the true zero). Future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations) and by cross-validating it with other more comprehensive measures.

@@

The dataset analysed during the current study are available in the GitHub repository \url{https://github.com/DominiqueMakowski/PHQ4R}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{PHQ-4, depression, anxiety, brief questionnaire, short scale\newline\indent Word count: 1158}
\usepackage{lineno}

@@

\abstract{%
The PHQ-4 is an ultra brief (4 items) screening questionnaire for depression and anxiety. We propose to add one aditional response option (``Once or twice'', in between ``Not at all'' and ``Several days'') to improve its sensitivity to milder alterations, and thus increase its usefulness in subclinical populations. Using Item Response Theory (IRT), we show that this new option does indeed capture specific portions of the measured constructs, and provide a scoring procedure that preserve the compatability with past versions.

\textbf{Significance Statement}. Developing reliable and sensitive instruments for mood disorders screening is critical in a global context marked by international crises (pandemics, wars), where increasingly more surveys are done online. In this study, we show that a small modification to the widely used PHQ-4 scale (adding the ``Once or twice'' response option) can increase its ability to capture the mild fluctuations prevalent in subclinical samples.
}

@@

The Patient Health Questionnaire-4 (PHQ-4) is an ultra brief measurement of core signs of depression and anxiety (Kroenke et al., 2009). It consists of two items for depression (PHQ--2, Kroenke et al., 2003) and anxiety (GAD--2, Kroenke et al., 2007), each corresponding to DSM-5 diagnostic symptoms for major depressive disorder (MDD) and generalized anxiety disorder (GAD). It has been validated across many languages and populations (Christodoulaki et al., 2022; Materu et al., 2020; Mendoza et al., 2022), becoming one of the most popular screening instruments for depression and anxiety (Maurer et al., 2018).

@@

\hypertarget{original-scale-and-revision}{%
\subsection{Original Scale and Revision}\label{original-scale-and-revision}}

@@

@@

The sample consists of 485 English-speaking participants (Mean age = 30.1, SD = 10.1, range: {[}18, 73{]}; Sex: 50.3\% females, 49.7\% males) from the general population who were administered the refined PHQ-4 online as part of another study, which data is available in open-access at \url{https://github.com/RealityBending/IllusionGameReliability}. This study was approved by the NTU Institutional Review Board (NTU IRB-2022-187). All participants provided their informed consent prior to participation and were incentivized after completing the study.

@@

\hypertarget{descriptive-statistics}{%
\subsection{Descriptive Statistics}\label{descriptive-statistics}}

@@

The objective of this study was to test the introduction of a ``Once or twice'' response option to the PHQ-4 to enhance its sensitivity to milder mood fluctuations. The fact that this new response option was the most prevalent response made by participants in our study is in itself evidence for its usefulness. IRT analysis further revealed that this response captures with precision a unique portion of anxiety and depression. Taken together, our results suggest that adding this option response increases the scale's potential to discriminate average mood levels (which are superior to zero) from lower-end extremes (the true zero).

The hypothesized added value in terms of finer sensitivity should be confirmed by cross-validating the PHQ-4R with more comprehensive measures of anxiety and depression. Additionally, evidence comparing the refined version against the original PHQ-4 would be useful to better understand whether the new ``Once or twice'' option pulls participants that would have otherwise answer ``Not at all'' or ``Several days''. Finally, future studies should further corroborate our statistical validation and test the usefulness of this revision on more targeted samples (e.g., clinical populations) to see if it affects the cut-offs proposed in the literature (and whether it allows for a better discrimination of mood disorders), and refine if need be the normalized scoring estimates in larger sample.

@@

The dataset analysed during the current study are available in the GitHub repository \url{https://github.com/DominiqueMakowski/PHQ4R}. We have created an R function for easy scoring according to the two methods described, which can be used as follows:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load the function}
\FunctionTok{source}\NormalTok{(}\StringTok{"https://raw.githubusercontent.com/DominiqueMakowski/PHQ4R/main/score\_PHQ4.R"}\NormalTok{)}

\CommentTok{\# Enter the responses}
\FunctionTok{score\_PHQ4}\NormalTok{(}\AttributeTok{A1 =} \StringTok{"Not at all"}\NormalTok{, }\AttributeTok{A2 =} \StringTok{"Once or twice"}\NormalTok{,}
           \AttributeTok{D1 =} \StringTok{"Once or twice"}\NormalTok{, }\AttributeTok{D2 =} \StringTok{"several days"}\NormalTok{,}
           \AttributeTok{method =} \StringTok{"basic"}\NormalTok{)}
\DocumentationTok{\#\#   A1  A2  D1 D2 Anxiety Depression}
\DocumentationTok{\#\# 1  0 0.5 0.5  1     0.5        1.5}
\end{Highlighting}
\end{Shaded}
